---
title: "HW3"
author: "Amy Pitts"
date: "10/10/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggridges)
library(p8105.datasets)
```


# Problem 1 

This problem uses the Instacart data. DO NOT include this dataset in your local data directory; instead, load the data from the p8105.datasets using:
```{r}
data("instacart")
```

The goal is to do some exploration of this dataset. To that end, write a short description of the dataset, noting the size and structure of the data, describing some key variables, and giving illstrative examples of observations. Then, do or answer the following (commenting on the results of each):



```{r}
head(instacart)
```

This dataset contains `r nrow(instacart)` rows and `r nrow(instacart)` columns. This data is all about instacart orders. 
Observations are at the level of items in orders by users of the insta cart app. There are user variables/ order variables, user ID, order ID, order data, and order hour. There are also item variables -- name, aisle, department, and some numeric codes. 

Also should contain the structure of the data. For example the aisle is realte dto the department. Also the aisle name is numberic and also is the same as the aisle id. 



- How many aisles are there, and which aisles are the most items ordered from?

```{r}
instacart %>%
  count(aisle) %>%
  arrange(desc(n))
```
The most things are coming from fresh vegetables and fresh friuts. 


- Make a plot that shows the number of items ordered in each aisle, limiting this to aisles with more than 10000 items ordered. Arrange aisles sensibly, and organize your plot so others can read it.

Lets make a plot 
```{r}
instacart %>%
  count(aisle) %>%
  filter(n > 10000) %>%
  mutate (
    aisle = factor(aisle),
    aisle = fct_reorder(aisle, n) #re-order isle according to n
  ) %>%
  ggplot(aes(x = aisle, y = n)) +
  geom_point() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) #needed to tilt the words 
```


- Make a table showing the three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”. Include the number of times each item is ordered in your table.

```{r}
instacart %>% 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) %>%
  group_by(aisle) %>%
  count(product_name) %>%
  mutate(rank = min_rank(desc(n))) %>%
  filter(rank < 4) %>%
  arrange(desc(n))# %>%
  #knitr::kable()
```



- Make a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week; format this table for human readers (i.e. produce a 2 x 7 table).

```{r}
instacart %>% 
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) %>%
  group_by(product_name, order_dow) %>%
  summarize( mean_hour = mean(order_hour_of_day)) %>%
  pivot_wider(
    names_from = order_dow,
    values_from = mean_hour
  )
```



# Problem 2

Accelerometers have become an appealing alternative to self-report techniques for studying physical activity in observational studies and clinical trials, largely because of their relative objectivity. During observation periods, the devices measure “activity counts” in a short period; one-minute intervals are common. Because accelerometers can be worn comfortably and unobtrusively, they produce around-the-clock observations.

This problem uses five weeks of accelerometer data collected on a 63 year-old male with BMI 25, who was admitted to the Advanced Cardiac Care Center of Columbia University Medical Center and diagnosed with congestive heart failure (CHF). The data can be downloaded here. In this spreadsheet, variables activity.* are the activity counts for each minute of a 24-hour day starting at midnight.

```{r data_q2}
activity = read_csv(file = "data/accel_data.csv") %>%
  janitor::clean_names()
```

- Load, tidy, and otherwise wrangle the data. Your final dataset should include all originally observed variables and values; have useful variable names; include a weekday vs weekend variable; and encode data with reasonable variable classes. Describe the resulting dataset (e.g. what variables exist, how many observations, etc).


(this is not tidy data) 
wide to long (so we need to make it longer)  pivot longer 
tweek other variables 
useful var names 
add a var to say weekend vs week day (mutate)
encode data with reasonable data classes, are activty count number? are days factor? 
is min of the day in the numberic format or in a factor. should be a numeric and need to change this 


- Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate accross minutes to create a total activity variable for each day, and create a table showing these totals. Are any trends apparent?


aggregate (group by and summarize problem ) 35 days and total activity count 
week number and day of the week to make it more readable 
(we need to use the factor var to change the order of the day of the week )



- Accelerometer data allows the inspection activity over the course of the day. Make a single-panel plot that shows the 24-hour activity time courses for each day and use color to indicate day of the week. Describe in words any patterns or conclusions you can make based on this graph.


plot just one 
24 hours of activity 
each min of each day -> (min x axis, active count y axis )
then we need to add in peices (goemetries and astetic colors) 
need a goem_line 
use color to indicate day of the week 
Describe 
(he is asleep in the night time and awake and moveing aorundi n the day time)








# Problem 3 
This problem uses the NY NOAA data. DO NOT include this dataset in your local data directory; instead, load the data from the p8105.datasets package using:

```{r}
data("ny_noaa")
```

The goal is to do some exploration of this dataset. To that end, write a short description of the dataset, noting the size and structure of the data, describing some key variables, and indicating the extent to which missing data is an issue. Then, do or answer the following (commenting on the results of each):

This data focuses on New York state from Jan 1, 1981 to Dec 31, 2010 and looks at the snowfall. 


- Do some data cleaning. Create separate variables for year, month, and day. Ensure observations for temperature, precipitation, and snowfall are given in reasonable units. For snowfall, what are the most commonly observed values? Why?

need to count the 


- Make a two-panel plot showing the average max temperature in January and in July in each station across years. Is there any observable / interpretable structure? Any outliers?

this is data manipulation steps and then plot them 
first step -> how do we need to organize the data (group by (station, month,year), and summarize(max temp), filter(just jan))
Do same thing for the july
use facating  

(can we see global warming happening)



- Make a two-panel plot showing (i) tmax vs tmin for the full dataset (note that a scatterplot may not be the best option); and (ii) make a plot showing the distribution of snowfall values greater than 0 and less than 100 separately by year.

These are two distinct plots and then we need to join them together 
probably patch work 
merge them to gether in some way 

first plot is a scatter plot, need to do a contor plot or a bin plot 

second need to filter, then use something like a box plots for each year 











